---
title: "Econometrics 3 - Problem Set 3"
author: "Pierre Pili - Isée Biglietti - Marie Gardie"
date: "16/02/2024"
output: pdf_document
indent: true
---

```{r, include = F, message = F}
# Libraries
library("tidyverse")
library("knitr")
library("stargazer")
library("haven")
library("descr")
library("broom")
library("AER")

# Imports
ppen97 <- read_dta("RAW/PPEN97.dta") %>%
  rename(score_1 = scglob) %>%
  mutate(anai = 1900 + anai)

psen95 <- read_dta("RAW/PSEN95.dta") %>%
  rename(
    score_f_6 = franel,
    score_m_6 = mathel,
    score_f_9 = brevfra1,
    score_m_9 = brevmat1,
    score_fl_9 = brevlv11
  ) %>%
  mutate(
    score_f_9 = 5 * score_f_9,
    score_m_9 = 5 * score_m_9,
    score_fl_9 = 5 * score_fl_9
  )
```

\section{1. Data Preparation}

### 1)

```{r, include = F, message = F}
# Function counting NAs
na_frequency <- function(x) {
  100 * mean(is.na(x))
}

summary97 <- ppen97 %>%
  select(score_1, score_f_3, score_m_3) %>%
  summarize(
    var = names(.),
    median = sapply(., median, na.rm = TRUE),
    mean = sapply(., mean, na.rm = TRUE),
    min = sapply(., min, na.rm = TRUE),
    max = sapply(., max, na.rm = TRUE),
    sd = sapply(., sd, na.rm = TRUE),
    `NAs (%)` = sapply(., na_frequency)
  )

summary95 <- psen95 %>%
  select(
    score_f_6, score_m_6, score_f_9,
    score_m_9, score_fl_9) %>%
  summarize(
    var = names(.),
    median = sapply(., median, na.rm = TRUE),
    mean = sapply(., mean, na.rm = TRUE),
    min = sapply(., min, na.rm = TRUE),
    max = sapply(., max, na.rm = TRUE),
    sd = sapply(., sd, na.rm = TRUE),
    `NAs (%)` = sapply(., na_frequency)
  )
```

```{r, echo = F}
kable(summary97, caption = "PPEN97 -- Scores in Years 1 and 3", digits = 1) #nolint
kable(summary95, caption = "PSEN95 -- Scores in Years 6 and 9", digits = 1) #nolint
```
As grades during year 9 were between 0 and 20 we rescaled them between 0 and 100 to make comparisons easier. The 1997 dataset (PPEN97) contains one variable for year 1 `score_1` and two for year 2 measuring French and Math test scores denoted `score_f_3` and `score_m_3`.
The 1995 dataset (PSEN95) also contains French and Math test scores for year 6 (`score_f_6`, `score_m_6`, `score_f_9`, `score_m_9`) and one additional variable for foreign language test scores `score_fl_9`.
We notice that year 9 contains more missing values with almost 40% for `score_f_9` and `score_m_9`.

### 2)

(see code)

```{r, echo = F}
ppen97 <- ppen97 %>%
  mutate(
    sexe = as.factor(sexe - 1),
    prior_area_1 = as.factor(prior_area_1 - 1),
    prior_area_3 = as.factor(prior_area_3 - 1),
    public_1 = as.factor(public_1 - 1),
    public_3 = as.factor(public_3 - 1),
    nati = as.factor(if_else(nati == 100, 1, 0)),
    age_1 = 12 * (1997 - anai) + 10 - mnai,
    age_3 = 12 * (yr_3 - anai) + 10 - mnai,
    relative_age = 12 - mnai
  )

psen95 <- psen95 %>%
  mutate(
    anai = 1900 + as.integer(substr(datenai, 5, 6)),
    mnai = as.integer(substr(datenai, 3, 4)),
    sexe = as.factor(as.integer(sexe) - 1),
    prior_area_6 = as.factor(prior_area_6 - 1),
    prior_area_9 = as.factor(prior_area_9 - 1),
    public_6 = as.factor(public_6 - 1),
    public_9 = as.factor(public_9 - 1),
    nati = as.factor(if_else(nateleve == 100, 1, 0)),
    age_6 = 12 * (yr_6 - anai) + 10 - mnai,
    age_9 = 12 * (yr_9 - anai) + 6 - mnai,
    relative_age = 12 - mnai
  )
```
\section{2. Naïve Estimation of Month Birth Effects}

### 1)

Firstly, pupil born in the beginning of the year will most likely join school a year sooner and conversely for those born at the end of the year. This is will reduce the actual age difference among each cohort. Moreover, students can be held back for a year or more which is more likely for students born at the end of the year. Both effects imply an underestimation of the true effect.
Secondly, there is a source of endogeneity as social economic backrounds can explain both the month of birth and the test score of the pupil. For instance, teacher's children are more generally born in may and are more sucessful at school. There is also a measurment problem with foreign born children who sometimes get assigned a date of birth, usually in January, which leads to January born children to be over-represented.
The OLS estimator will thus be biased.

### 2)

